Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-24 manuscript/response_to_reviews.Rmd

**bold - response**

~~~~

Dear Editor:

We have revised our manuscript, “Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest” (previously “Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest”), according to the recommendations of reviewers. This has resulted in an easier to read manuscript with more robust support.

We addressed all of the reviewers’ points, as detailed below, resulting in several major changes:

* At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric of drought resistance (using an ARIMA model to project what growth would have been in drought year based on the trend over the past 10 years). Results were similar, and conclusions the same. We now present those results as a secondary model (mostly in SI).

* We dropped crown position as a factor in the statistical models of drought resistance, and instead include it as part of an enhanced focus on the differences in microenviroment experienced by trees of different height (now in Table 1; Fig. 2). This change was prompted by comments by two reviewers (R1, R3) that canopy position could not be reliably projected back to 1966 or disentangled from height.

* We shifted to more figures and fewer tables in the main article (as requested by R1, R3, and R4). We added two figures to the main text: one showing Rt differences across species (Fig. 3) and one visualizing the model results (Fig. 4). We moved former tables 4 and 5 to the SI. We also added several new figures and tables to the SI to provide information requested by the reviewers.

We hope that this manuscript will now be found suitable for publication in New Phytologist, and we look forward to your response.

Best regards,

Kristina Anderson-Teixeira (on behalf of all authors)

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
Updated with ian final pass #117

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA. 
First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al. (2015).
We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance.

@@

Species' trait data were collected in August 2018 (Tables 2-3; Fig. **S4**). We sampled small, sun-exposed branches up to eight meters above the ground from three individuals of each species in and around the ForestGEO plot. Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets under opaque plastic bags before measurements were taken. Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves. Wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to fresh volume, which was estimated using Archimedes’ displacement. We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4 mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].

@@

For each drought year, we calculated a metric drought resistance ($Rt$) as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011]. Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively. Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$). The two metrics were strongly correlated (Fig. **S5**). Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info. In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of [@desoto_low_2020] suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.

@@

We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$ln[TWI]$ and species traits as specified above.
We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1). 
If the variable appeared in some but not all of these models, and its sign was consistent across models, we considered this partial support/rejection. 
In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.

@@

Species, as a factor in ANOVA, had significant influence (p<0.05) on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig. **S4**).
Drought resistance also varied across species, overall and in each drought year (Fig. **3**). Significant differences in $Rt$ across species were most pronounced in 1966 with a total of seven distinct groupings, while 1977 had four and 1999 had two.
Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).

>>>>

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA. 
First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance.

@@

Species' trait data were collected in August 2018 (Tables 2-3; Fig. **S4**). We sampled small, sun-exposed branches up to eight meters above the ground from three individuals of each species in and around the ForestGEO plot. Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets under opaque plastic bags before measurements were taken. Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves. Wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to fresh volume, which was estimated using Archimedes’ displacement. We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4 mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapor pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].

@@

For each drought year, we calculated a metric drought resistance ($Rt$) as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011]. Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively. Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$). The two metrics were strongly correlated (Fig. **S5**). Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info. In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of @desoto_low_2020 suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.

@@

We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$ln[TWI]$ and species traits as specified above.
We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1). 
If the variable appeared in some but not all of these models, and its sign was consistent across models, we considered this partial support/rejection. 
In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model. Visualization of the best mixed effects model per drought scenario (Fig. **4**) was created by the visreg package [@R-visreg].

@@

Species, as a factor in ANOVA, had significant influence (p<0.05) on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig. **S4** as characterized by the agricolae package @R-agricolae).
Drought resistance also varied across species, overall and in each drought year (Fig. **3**). Significant differences in $Rt$ across species were most pronounced in 1966 with a total of seven distinct groupings, while 1977 had four and 1999 had two.
Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
